%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% LaTeX Example: Project Report
%
% Source: http://www.howtotex.com
%
% Feel free to distribute this example, but please keep the referral
% to howtotex.com
% Date: March 2011 
% 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% How to use writeLaTeX: 
%
% You edit the source code here on the left, and the preview on the
% right shows you the result within a few seconds.
%
% Bookmark this page and share the URL with your co-authors. They can
% edit at the same time!
%
% You can upload figures, bibliographies, custom classes and
% styles using the files menu.
%
% If you're new to LaTeX, the wikibook is a great place to start:
% http://en.wikibooks.org/wiki/LaTeX
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Edit the title below to update the display in My Documents
%\title{Project Report}
%
%%% Preamble
\documentclass[paper=a4, fontsize=11pt]{scrartcl}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{fourier}

\usepackage[english]{babel}                                                         % English language/hyphenation
\usepackage[protrusion=true,expansion=true]{microtype}  
\usepackage{amsmath,amsfonts,amsthm} % Math packages
\usepackage[pdftex]{graphicx}   
\usepackage{url}
\usepackage{listings}
\usepackage{arydshln}

%%% Custom sectioning
\usepackage{sectsty}
\allsectionsfont{\normalfont\bfseries}


%%% Custom headers/footers (fancyhdr package)
\usepackage{fancyhdr}
\pagestyle{fancyplain}
\fancyhead{}                                            % No page header
\fancyfoot[L]{}                                         % Empty 
\fancyfoot[C]{}                                         % Empty
\fancyfoot[R]{\thepage}                                 % Pagenumbering
\renewcommand{\headrulewidth}{0pt}          % Remove header underlines
\renewcommand{\footrulewidth}{0pt}              % Remove footer underlines
\setlength{\headheight}{13.6pt}

\newtheorem{defn}{Definition}[section]

%%% Equation and float numbering
\numberwithin{equation}{section}        % Equationnumbering: section.eq#
\numberwithin{figure}{section}          % Figurenumbering: section.fig#
\numberwithin{table}{section}               % Tablenumbering: section.tab#


%%% Maketitle metadata
\newcommand{\horrule}[1]{\rule{\linewidth}{#1}}     % Horizontal rule
%\DeclareMathOperator{\dim}{dim}

\title{
        %\vspace{-1in}  
        \usefont{OT1}{bch}{b}{n}
        \normalfont \normalsize \textsc{University of Strasbourg} \\ [25pt]
        \horrule{0.5pt} \\[0.4cm]
        \huge Intermediary Internship Report \\
        \horrule{2pt} \\[0.5cm]
}
\author{
        \normalfont                                 \normalsize
        Thomas Kuntz \\                                \normalsize
        Master RISE \\                                \normalsize
        UFR d'Informatique \\                                \normalsize
        Supervisor : Cédric Bastoul\\
}
\date{}


%%% Begin document
\begin{document}
\maketitle
\thispagestyle{empty}

\clearpage

\tableofcontents
\clearpage

\section{Presentation of the laboratory and research team}
This intership falls within the research conducted in the ICPS team (Parallel and Scientific
Computing) of the ICube laboratory at the university of Strasbourg, and the CAMUS group
(Compilation for MultiCore Architectures) at Inria, the French Institute for Research
in Computer Science. ICPS/CAMUS focuses at providing developers with theoretical and software
tools to develop efficient applications for parallel architectures without sacrificing
productivity.

Parallel systems are now omnipresent, from supercomputers to mobile devices,
and their effective use requires developers to design parallel programs or to rewrite
legacy sequential applications. However, parallel programming is still complex and out
of reach for non experts, at either designing, writing or debugging level. To address
this issue, ICPS/CAMUS is developing source-to-source compiler techniques for automatic
program optimization and parallelization. Using these technologies, developers can
continue to write sequential programs while the mapping to parallel architecture is
computed automatically.

The team has 7 faculty members and a dozen of PhD students,
postdocs or engineers. Its members are deeply involved in European and international
collaborations on scientific projects both with academia and industry.
Its most renowned technical contributions in this field include
the PolyLib library, the code generator CLooG, the dynamic optimizer VMAD or the parallelizer
of binary programs BinPar.

\section{Internship mission : High-Level Optimization Driven by Statement Profiling}
Many computation-intensive programs spend a very large amount of their execution
time inside nested loops. Thus, loop nest optimization is one of the major approach used
for program optimisation.

To represent the loop nest, you can use the Polyhedral Model, a powerful mathematical model
that allows to express loop nest as union of polytopes and loop optimizations as
transformation of these polytopes.

The game is to find the "good" polytope transformations to end up with
transformed polytopes representing loop nest (which doesn't violate dependences) 
that can be parallelized, vectorized, tiled, etc. with better properties than the original
polytope.

The Polyhedral Model is really good with short and simple codes, but struggle with
large and complex codes.The reason is twofold. First, polyhedral techniques exhibit
a very high worst-case complexity which hampers its applicability to large codes.
Second, contradicting optimization objectives may result in suboptimal global
decisions


We would like to address these issues by proposing a new technique : before any
high-level optimizations are used on the source code, a first pass is applied (our pass),
aggregating "near behaving" statements together using their profile.
"Near behaving" statements describe statements that have similar properties,
like data reuse/locality, parallelism, vectorization etc. that constitute their profile.
By doing that, we hope to ease the work of the high-level optimizer by "reducing" the size
of large and complex code, and prevent it from separating statements that should
not be separated (for example if they have a lot of data reuse in common).

\section{Work done}
    \subsection{Learning about the polyhedral model}
        The polyhedral model is (as explained in \cite{Bas'12}) a mathematical model that allows to represent program
        part as union of polytopes and affine sets. Program parts that can fit the model
        are called \textit{static control parts} or \textit{SCoPs} for short.
        Scops are generally loop-based, and need to have affine loop bounds, conditions and
        array subscripts to fit the polyhedral model.

        The polyhedral model consist of three mathematical objects based on unions of relations :
        \begin{itemize}
            \item The \textit{Iteration Domain}, that provide the set of all the iteration scanned
                %parcouru
                by a specific statement, using affine constraints. It's these affine constraints 
                that can be seen as a $\mathbb{Z}$-polytope.
            \item The \textit{Scattering Relations} (or \textit{Scheduling Relations},
                or even\textit{Mapping Relations}), that give us the ordering which
                all the statement instances will be executed.
            \item The \textit{Access Relations} that model read/write accesses made by
                the different arrays and simple variables.
        \end{itemize}

            \subsubsection{Iteration Domain}
                To understand Iteration Domain, you must first understand what a \textit{statement instance}
                is. A statement instance is one particular execution of a statement.
                When a statement is located inside a loop nest, it can be associated to the value
                of the outer-loop counters (called \textit{iterators}.
                The Iteration Domain will be the set of the iterators of all the statement
                instance.

                Because the outer loop bound can be affine or parameterized, it is generally impossible to
                know their actual value at compilation time. Even if we knew the value of the
                bounds, the number of iterators could be to enormous to count them all into sets.
                That's why the sets are expressed as a system of affine constraints.
                These affine constraints define a $\mathbb{Z}$-polyhedron, hence the name of the model.
                The relation used to represent the Iteration Domain is :
                \begin{center}
                    $ \mathcal{D}_S(\vec{p}) = \left\{() \to \vec{\imath_S} \in \mathbb{Z}^{\dim(\vec{\imath_S})}
                    \middle|
                    \left[D_S\right]\begin{pmatrix}\vec{\imath} \\ \vec{p} \\ 1\end{pmatrix}
                    \geq \vec{0}
                    \right\}$
                \end{center}
                where $\vec{\imath}$ is the $\dim(\vec{imath})$-dimensional iteration vector and
                $D_S \in \mathbb{Z}^{m_{\mathcal{D}_S} \times (\dim(\vec{\imath})+\dim(\vec{p})+1)}$
                is an integer matrix where $m_{\mathcal{D}_S}$ is the number of constraints.
                \\

        \begin{lstlisting}[frame=single, language=C, caption={Simple code for polyhedral model example}, label={lst:polyhedral_example}]
                for(i=0 ; i<2*N - 1 ; i++)
        S1:         A[i] = 42;
                for(i=0 ; i<N ; i++)
                    for(j=0 ; j<N ; j++)
        S2:             B[i][j] += 42;
        \end{lstlisting}
                
                For example, in Listing~\ref{lst:polyhedral_example}, we have for
                S1 and S2 the iteration domain :
                \begin{itemize}
                    \item[]$ \mathcal{D}_{S1}(N) = \left\{() \to \begin{bmatrix}i\end{bmatrix} \in \mathbb{Z} \middle|
                    \left[\begin{array}{c:c:c}
                            1 & 0 & 0 \\
                            -1 & 2 & -2
                    \end{array}\right]
                    \left(\begin{array}{c}
                            i \\ \hdashline
                            N \\ \hdashline
                            1 
                    \end{array}\right)
                    \geq \vec{0}
                    \right\} $,
                
                    \item[]$ \mathcal{D}_{S2}(N) = \left\{() \to \begin{bmatrix}i\end{bmatrix} \in \mathbb{Z} \middle|
                    \left[\begin{array}{c:c:c:c}
                            1 & 0 & 0 & 0\\
                            -1 & 0 & 1 & -1\\
                            0 & 1 & 0 & 0\\
                            0 & -1 & 1 & -1
                    \end{array}\right]
                    \left(\begin{array}{c}
                            i \\
                            j \\ \hdashline
                            N \\ \hdashline
                            1 
                    \end{array}\right)
                    \geq \vec{0}
                    \right\} $.
                \end{itemize}

                In the case of \textit{if-then-else} enclosed statements, the iteration domain
                for the statement will be an union of disjoint polytope, each representing a part
                of the \textit{if-then-else} condition.

            \subsubsection{Scattering Relations}
                Defining the domain of iteration is great, but it doesn't give us any information
                on the order in which will be executed each statement instance.

                There is different type of ordering relations :
                \begin{description}
                    \item[Scheduling] is about ordering the statement instances in time by
                        giving them a logical date vector indicating when each will be executed.
                        The data vector is generally read in \textit{lexicographically} order, for example
                        $(1,2,3)$ is executed before $(1,2,4)$ but after $(1,1,4)$.
                    \item[Allocation] or \textbf{placement} is about ordering statement instances
                        in space by assigning to each statement instances a logical stamp
                        indicating the processor number in which it will be executed.
                        Statement instances not sharing the same logical stamp can be
                        executed in parallel in different processor.
                \end{description}

                It is possible to express each ordering relations as separate relation,
                but it is also possible to aggregating them together : the result would be
                a relation between the iterators set and a set of ordering vector which
                would have some dimensions allocated to time ordering (scheduling) and
                some other to space ordering (allocation).

                For a statement $S$, the scattering relation can be written as follow :
                \begin{center}
                    $\theta_S(\vec{p}) = \left\{\vec{\imath_S} \to \vec{t_S} \in \mathbb{Z}^{\dim(\vec{\imath_S})}\times\mathbb{Z}^{\dim(\vec{t_S})}
                    \middle|
                    \left[T_S\right]\begin{pmatrix}\vec{t_S}\\ \vec{\imath} \\ \vec{p} \\ 1\end{pmatrix}
                    \geq \vec{0}
                    \right\}$,
                \end{center}
                where $\vec{\imath}$ is the $\dim(\vec{\imath})$-dimensional iteration vector
                and $T_S \in \mathbb{Z}^{m_{\theta_S}\times(\dim(\vec{\imath_S})+\dim(\vec{t_S})+\dim(\vec{p})+1)}$
                is an integer matrix where $m_{\theta_S}$ is the number of constraints.

                For example, for Listing~\ref{lst:polyhedral_example}, we have for S1 and S2
                statement the original scattering of the program :
                \begin{itemize}
                    \item[]
                        $\theta_{S1}(N) = \left\{(i) \to \begin{pmatrix}t^{1}_{S1}\\t^{2}_{S1}\\t^{3}_{S1}\end{pmatrix} \in
                            \mathbb{Z}\times\mathbb{Z}^{3}
                            \middle|
                            \left[\begin{array}{ccc:c:c:c}
                                    -1 & 0 & 0 & 0 & 0 & 0 \\
                                    0 & -1 & 0 & 1 & 0 & 0 \\ 
                                    0 & 0 & -1 & 0 & 0 & 0
                            \end{array}\right]
                            \left(\begin{array}{c}
                                t^{1}_{S1} \\
                                t^{2}_{S1} \\
                                t^{3}_{S1} \\ \hdashline
                                i \\ \hdashline
                                N \\ \hdashline
                                1
                            \end{array}\right)
                            = \vec{0}
                            \right\}$,\\
                            which can be simplify to $\theta_{S1}(N)(i)=\begin{pmatrix}0\\i\\0\end{pmatrix}$.
                    \item[]
                        $\theta_{S2}(N) = \left\{(i) \to \begin{pmatrix}t^{1}_{S2}\\t^{2}_{S2}\\t^{3}_{S2}\\t^{4}_{S2}\\t^{5}_{S2}\end{pmatrix} \in
                            \mathbb{Z}\times\mathbb{Z}^{3}
                            \middle|
                            \left[\begin{array}{ccccc:cc:c:c}
                                    -1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1\\
                                    0 & -1 & 0 & 0 & 0 & 1 & 0 & 0 & 0\\ 
                                    0 & 0 & -1 & 0 & 0 & 0 & 0 & 0 & 0\\
                                    0 & 0 & 0 & -1 & 0 & 0 & 1 & 0 & 0\\ 
                                    0 & 0 & 0 & 0 & -1 & 0 & 0 & 0 & 0
                            \end{array}\right]
                            \left(\begin{array}{c}
                                t^{1}_{S2} \\
                                t^{2}_{S2} \\
                                t^{3}_{S2} \\
                                t^{4}_{S2} \\
                                t^{5}_{S2} \\ \hdashline
                                i \\ \hdashline
                                N \\ \hdashline
                                1
                            \end{array}\right)
                            = \vec{0}
                            \right\}$,\\
                            which can be simplify to $\theta_{S2}(N)\begin{pmatrix}i\\j\end{pmatrix}=\begin{pmatrix}0\\i\\0\\j\\0\end{pmatrix}$.
                \end{itemize}
                The first dimension for S1 is 0 and for S2 is 1 : it is to ensure that S2
                is always executed after S1.
                In our example, it doesn't matter because S1 and S2 doesn't have any dependence
                whatsoever, and they could be run in parallel without problem, but for the sake
                of the example, we've imposed that S1 is executed before S2.

            \subsubsection{Access Relations}
                The \textit{Access Relations} models the memory access of the different
                array references in a statement. It map the statement instances to memory locations
                for every array reference in the statement. Non-array reference can be seen as
                0-dimensional array.

                The access relation can be written in the general form :
                \begin{center}
                    $\mathcal{A}_{S,r}(\vec{p}) = 
                    \left\{
                        \vec{\imath_S} \to \vec{a}_{S,r} \in \mathbb{Z}^{\dim(\vec{\imath_S})}\times\mathbb{Z}^{\dim(\vec{a}_{S,r})}
                        \middle|
                        \left[A_{S,r}\right]\left(\begin{array}{c}\vec{a}_{S,r}\\\vec{\imath_S}\\\vec{p}\\1\end{array}\right)
                        \geq \vec{0}
                    \right\}$
                \end{center}
                where 
                \begin{itemize}
                    \item $r$ is the number of the reference in the statement (every references are
                        numbered, from left to right generally),
                    \item $\vec{\imath}$ is the $\dim(\vec{\imath})$-dimensional iteration vector
                    \item $T_S \in \mathbb{Z}^{m_{\theta_S}\times(\dim(\vec{\imath_S})+\dim(\vec{t_S})+\dim(\vec{p})+1)}$
                        is an integer matrix where $m_{\theta_S}$ is the number of constraints.
                \end{itemize}

                The read or write nature of the memory access is not expressed in this form,
                and need to be given (for example, to compute dependences).

            \subsubsection{OpenScop Framework}
                OpenScop~\cite{openscop} is an open specification that define a file format
                and a set of data structures to represent a static control part (SCoP).

                OpenScop format is intended to be a stable bridge between polyhedral tools
                and optimizers allowing interchangeable tools and flexible toolchains.

                A library named \textit{osl} is provided to read/write OpenScop files
                and manipulate the different data structures the specification define.
                The file format as well as the library has been designed to accept extensions.

                Substrate, the tool performing our optimization pass uses OpenScop as input and output
                format.
    \subsection{Data reuse}
        \subsubsection{Difference between Reuse and Locality}
            It's important to differentiate \textit{reuse} from \textit{locality}.
            We say that there is \textit{reuse} for a data when this data is
            access multiple time during the execution of a loop nest, during
            the same iteration or not. It can be from different references or different statements.
            
            When there is reuse, it's possible (but not guaranteed) to have
            \textit{locality}. We say that there is \textit{locality} when the
            reused data remains in memory hierarchy level targeted 
            (eg : the CPU cache in our case) between 2 reuses.
            For that to be the case, the amount of other data accessed between the
            2 reuses must not exceed the cache size, or else the reused data will be
            evicted from the cache, causing a cache miss and forcing the cache
            to fetch the data from the RAM (and RAM have higher latency compared to CPU cache).


            Thus, we can say that the \textit{reuse} is linked to the way indexing
            functions (for array references) are written, whereas \textit{locality} is
            inherent to the iterations order/the way the loops are written.

        \subsubsection{Types of Reuse}
            
%\begin{lstlisting}[frame=single, language=C, caption=Reuse example, label={lst:reuse_example}]
%for(i=0; i<N ; i++)
%    for(j=0; j<N ; j++)
%    {
%        A[i][j] = B[i] + B[i+1];
%        C[i][j] = B[i+2] * 2;
%    }
%\end{lstlisting}

            \paragraph{Self-Temporal}
                Self-Temporal reuse occurs when the same reference access the same
                data location.
                
%                For example, in Listing \ref{lst:reuse_example}, the reference \verb'B[i]' carry Self-Temporal
%                reuse :\verb'B[0]',\verb'B[1]',\verb'B[2]' ... \verb'B[N-1]' are accessed
%                \textit{j} times by the same reference \verb'B[i]'.
%                And for the same reason, \verb'B[i+1]' and \verb'B[i+2]' also carry Self-Temporal reuse.

            \paragraph{Self-Spatial}
                Self-Spatial reuse occurs when the same reference access the same
                line in an array (when in row-major). Especially, the space between
                two accessed data location should not be greater than a cache line size,
                or else there won't be any spatial locality.

%                For example, in Listing \ref{lst:reuse_example}, the reference \verb'A[i][j]' carry
%                Self-Spatial reuse.

                Self-Temporal reuse can be considered as a special case of
                Self-Spatial reuse : accessing the same data location mean that
                the same array line is also accessed.
            \paragraph{Group-Temporal}
                Group-Temporal reuse occurs when different references access the same
                data location. It can occur inside a same statement or across different
                statements.

%                For example, in Listing \ref{lst:reuse_example}, the reference
%                \verb'B[i]', \verb'B[i+1]' and \verb'B[i+2]' carry Group-Temporal :
%                the data accessed by \verb'B[i+2]' when $i=M$ will be access by
%                \verb'B[i+1]' when $i=M+1$ and by \verb'B[i]' when $i=M+2$.

            \paragraph{Group-Spatial}
                Group-Spatial reuse occurs when different references access the same
                line in an array (when in row-major). Again, the space between
                two accessed data location should not be greater than a cache line size,
                or else there won't be any spatial locality.
                
%                For example, in Listing \ref{lst:reuse_example}, the reference
%                \verb'B[i]', \verb'B[i+1]' and \verb'B[i+2]' carry Group-Spatial :
%                they all access to different data location, but on the same line.\\
%                \\

                So as you can see, a single reference can carry multiple type of reuse.
                A statement being generally composed of multiple references, it 
                can be hard to automatically favor some statements instead of others.
    \subsection{How to profile data reuse}
        \subsubsection{Reuse Vector Space}
            The notion of Reuse Vector Space come from Wolf et. al.\cite{Wolf'91} and describe
            the vector subspace of the Iteration vector space where the reuse of
            a reference can be taken advantage to induce locality.
            There is as many reuse vector space as there is type of reuse.

            Wolf et. al. use the concept of uniformly generated references proposed
            by Gannon et. al.\cite{Gannon:1988:SCL:50454.50460} for estimating reference
            windows :
            \begin{defn}
            \label{sec:uniformly_generated}
                Let $n$ be the depth of a loop nest, and $d$ be the dimensionality of
                an array $A$. Two references $A[\vec{f}(\vec{\imath})]$ and
                $A[\vec{g}(\vec{\imath})]$, where $\vec{f}$ and $\vec{g}$ are indexing
                functions $Z^{n} \rightarrow Z^{n}$, are called \textbf{uniformly generated} if :
                \begin{center}
                    $\vec{f}(\vec{\imath}) = H\vec{\imath}+\vec{c}_f$ and $\vec{g}(\vec{\imath}) = H\vec{\imath}+\vec{c}_g$
                \end{center}
                where $H$ is a linear transformation and $\vec{c}_f$ and $\vec{c}_g$ are constant vectors.
            \end{defn}
            
            We can partition references into equivalence class, called \textit{uniformly generated sets},
            when the references operate on the same array, and have the same $H$.
            You can see $H$ as a part of one of the access function (in the Polyhedral Model representation) of this statement :
            it's the access function matrix without the part handling the constant vector.

            The problem with that method, is that it would put the for example the references
            $C[i][j],C[i][j+1],C[i-1][j]$ in the same class : they all share the
            same $H$, but they don't have the same reuse space at all.
            For example : $C[i][j],C[i-1][j]$ exhibit group-spatial reuse in the $i$ dimension,
            whereas $C[i][j],C[i][j+1]$ exhibit group-spatial reuse in the $j$ dimension. So they
            should not be group together in the same class.
            \medskip

            So the definition of the \textit{uniformly generated set} is not restrictive enough,
            and in our case, we propose an addition to it : for two references $A$ and
            $B$ (as expressed in Definition~\ref{sec:uniformly_generated}) to be in the same class :
            \begin{itemize}
                \item They need to have the same $H$.
                \item with $\vec{c}=\vec{c_g} - \vec{c_f}$, $\vec{c}$ need to have $0$
                    for all the elements of the vector \textbf{except one}.
            \end{itemize}

            The reuse spaces would then be computed for each type of reuse
            with the methods described in~\cite{Wolf'91} : statements where
            array references share the same reuse space should be aggregated (if
            it doesn't violate any data dependence).

        \subsubsection{The Simple Algorithm}
            The \textit{Simple Algorithm}~\cite{Kennedy94maximizingloop} is a
            optimization algorithm that improve data locality via Loop Fusion.
            The idea is to fusion vertices of a directed graph when these vertices
            are connected by non-fusion-preventing arcs.
            In this graph, the vertices represent the different loop nests of a program part,
            and the arcs represent data dependences between statements of different loop nests.

            If the dependence is fusion-preventing~\cite{Bacon:1994:CTH:197405.197406} or if the loop headers are incompatible,
            the arc representing the dependence will be marked fusion-preventing.

            The algorithm is a modification of the greedy algorithm that partition
            the nodes of the graph without sequentializing a parallel node/loop.
            This algorithm doesn't give us any metric or way to evaluate a statement
            reuse, so it doesn't help use for the reuse evaluation side, but the idea
            of a dependence graph based greedy algorithm is interesting : with statement
            instead of loop nest being the nodes, this kind of greedy algorithm 
            could fusion statements that are inside different loop nest in the SCoP.
            It means we would also need to redefine the \textbf{Safety} and \textbf{Profitability}
            notions of the algorithm.

        \subsubsection{Data and Loop Transformations}
            An algorithm that transforms a loop nest to improve cache locality
            by a combination of loop and data transformations is presented in~\cite{Kandemir99improvingcache}.

            There's not much to say about it. The goal of the algorithm is to
            compute a transformation matrix and array layout (and the explanation is a little
            vague), so it isn't applicable in our case and it doesn't give us any metric or way to evaluate the reuse.
        
        \subsubsection{LoopCost Algorithms}
            The LoopCost algorithm~\cite{McKinley:1996:IDL:233561.233564} provide
            us with a cost model that allow to guide some loop transformation like
            loop permutation, loop reversal, loop fusion, loop distribution,
            by comparing the cost with and without the transformation.
            
            The LoopCost algorithm work like this : 
            Each loop $l$ in a loop nest is considered as a candidate for the innermost
            position in the nest. For each loop $l$, LoopCost compute and sum the number
            of cache line access by a group of references (the groups are formed
            using the RefGroup algorithm) when $l$ is placed as the innermost loop.

            As a result, you have a cost for each loop when they are considered as innermost.
            This cost enable to possibility to compare with hard numbers
            a loop nest before and after some loop transformations, which allow to discard
            or apply the transformations if the new cost is worse or better than the
            original cost.
            \bigskip

            Because the LoopCost is loop transformation oriented and because it doesn't
            use the number of references in a reference group in the computation
            of the cost, it is not really applicable at our situation.
        \subsubsection{Automatic Polyhedral Parallelization and Locality Optimization Algorithm}
            Let $S_i$ and $S_j$ two statement of a program with a data dependence $e$
            from $S_i$ to $S_j$.\\
            Let $\vec{s}$ and $\vec{t}$ the iterators of respectively $S_i$ and $S_j$.\\
            Let $\theta_{S_i}$ and $\theta_{S_j}$ the scattering function of respectively $S_i$
            and $S_j$.\\
            Consider the following affine $\delta_e$ :
            \begin{center}
                $\delta_e\left(\vec{s},\vec{t}\right) = \theta_{S_j}\left(\vec{t}\right) - \theta_{S_i}\left(\vec{s}\right)$
            \end{center}

            According to the paper~\cite{Bondhugula:2008:PAP:1379022.1375595},
            the affine form $\delta_e\left(\vec{s},\vec{t}\right)$ is very significant
            because it gives us ``the number of hyperplanes the dependence $e$ traverses
            along the hyperplane normal $\theta$. If $\theta$ is used as a space loop to generate tiles
            for parallelization, this function is a factor in the communication
            volume. On the other hand, if $\theta$ is used as a sequential loop, it
            gives us a measure of the reuse distance.''

            If there is a upper bound to $\delta_e\left(\vec{s},\vec{t}\right)$,
            minimizing it (or even nullify it), would be the same as minimizing (or nullify)
            the number of hyperplanes that would be communicated as a result of
            the dependence at the tile boundaries.

            The problem is that minimizing $\delta_e\left(\vec{s},\vec{t}\right)$
            can be really difficult because it ends up in an objective non-linear
            in loop variables and hyperplane coefficients.

            We can avoid this problem if we use a bounding function $v(\vec{p}) = u.\vec{p} + w$ such that :
            \begin{center}
                $\delta_e\left(\vec{s},\vec{t}\right) \leq v(\vec{p})$
            \end{center}
            i.e.,
            \begin{center}
                $v(\vec{p}) - \delta_e\left(\vec{s},\vec{t}\right) \geq 0$
            \end{center}

            With this last inequality, we can apply the Farkas Lemma :
            \begin{center}
                $v(\vec{p}) - \delta_e\left(\vec{s},\vec{t}\right) \equiv \lambda_{e0}
                + \sum\limits_{k=1}^{m_e}{\lambda_{ek}\mathcal{P}_{e}^k},\qquad \lambda_{ek} \geq 0$
            \end{center}
            where $P_e^k$ is a face of $P_e$.

            From this new form, you can get a linear equality and inequality system,
            which can then be handled by PIP software to find minimal solution to it.

            This algorithm doesn't give us any metric or way to evaluate reuse, so
            we can't use it.

    \subsection{SuBStrAte : the Similar Behaving Statement Aggregator}
        Substrate is the program that implements what we learned during the internship.
        Given an input Openscop file, it :
        \begin{itemize}
            \item Analyze the properties of the different statement of the scops
                contained in the file : data reuse, vectorization, parallelization, etc...
                using the metrics found in the literature.
            \item Optimize the scops by aggregating statements when legal and preferable.
                To decide if it is preferable, we rate how much similarly two statements
                behave, for example we rate their common data reuse. The result is a rating,
                between 0 and 1. If this rating is greater than a specific threshold, we
                aggregate the two statements.
            \item Produce an output Openscop file containing the optimized scops.
        \end{itemize}

        For now, the analyze and optimization only care about data reuse. The data reuse
        profile of each statement is constructed by :
        \begin{itemize}
            \item All references in a statement are grouped by the name/id.
            \item All the reference of a same id are then grouped by Uniformly Generated Set
                (when they share the same H matrix).
                \item All the references in an Uniformly Generated Set are finally grouped
                    by equivalence class.
        \end{itemize}
        We compute the data reuse rating between two statements by counting the number of
        references that share the same name, H matrix and equivalence class between the
        two statements. We divide this number by the total number of references in the two
        statements. To be aggregated, two statements need to have a rating greater than a
        specific threshold, given by the user (a default one is given but it has been chosen
        a little randomly).

        For now, the aggregating algorithm only try to aggregating successive statements,
        starting from the first two and then iterate until the last ones.

\section{Work left}
    \subsection{Learning about other properties and their metrics}
        We've started the work on SuBStrate by learning about and implementing profiling of
        data reuse. There's a lot more kind of properties that can be used to profile
        and aggregate statements, like vectorization, parallelization, tiling hyperplane, etc...

        The next in line is parallelization. We have to read papers about how we can profile
        it in a statement, what metrics could be used for that, then trying to implement
        it and find a way to rate it with a number between 0 and 1, test it, etc...
    \subsection{Improve SuBStrAte aggregation and rating algorithms}
        For now, the aggregation algorithm (successive statement only, from the first two,
        to the last ones) is a little simple, and there could be other ways, better ways
        to do it. We thought about maybe using a greedy algorithm to prioritize successive
        statements that have the best rating, instead of start from the first two statements.
        We also thought about maybe a greedy algorithm, but used in a graph, where nodes
        are the statement, so that an aggregation can take place between statements that
        are not successive (with the aggregation still being legal).


\bibliographystyle{plain}
\bibliography{biblio}
\end{document}
